A Responsible Machine Learning Workflow\\\normalsize{With Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2020-01-31 article-information-2019.tex
improvements to XNN section

The simulated data (see Subsection \ref{ssec:sim_data}) is based on the well-known Friedman datasets. Its known feature importance and discrimination characteristics are used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree}, \cite{friedman1991multivariate}. The mortgage data (see Subsection \ref{ssec:mort_data}) is sourced from the Home Mortgage Disclosure Act (HMDA) database and is a fairly realistic data source for demonstrating the template workflow.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. Because the unconstrained ML models, gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), do not exhibit convincing accuracy benefits on the simulated or mortgage data and can also present previously delineated risks, further explanation and discrimination analysis are applied only to the constrained, interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} (See subsection \ref{ssec:mgbm}) and XNNs (see Subsection \ref{ssec:xnn}, \cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analyses. MGBM and XNN interpretable model architectures are selected for this text because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated to also evaluate the constrained versions of these models.

The same can be said of the selected explanation methods and discrimination tests. Due to their post-hoc nature, they can often be shoe-horned into existing ML workflows and pipelines. Presented explanation techniques include PD and ICE (see Subsection \ref{ssec:pd_ice}) and Shapley values (see Subsection \ref{ssec:shap}) \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discussed discrimination testing methods (see Subsection \ref{ssec:di}) include AIR, ME, SMD, and confusion matrix metrics \cite{feldman2015certifying}, \cite{cohen1988statistical}, \cite{cohen1992power}.\footnote{Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix metrics are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available through GitHub (see Subsection \ref{ssec:soft}).

@@

Simulated data is created based on a function first proposed in \citet{friedman1979tree} and extended in \citet{friedman1991multivariate}:

@@

\noindent where $\mathbf{X}_{\text{Friedman},j}$ are random uniform features in $[0,1]$. In Friedman's texts, a Gaussian noise term was added to create a continuous output feature for testing spline regression methodologies. In this manuscript, the signal generating function and input features are modified in several ways.  Two binary features, a categorical feature with five discrete levels, and a bias term are introduced into $f$ to add a degree of complexity that may more closely mimic real-world settings. 
%When generating the data, random draws for each observation are constructed such that the features can be correlated with each other, some highly so. For example, X1 and X3 have a Pearson correlation coefficient equal to 0.58.
For binary classification analysis, the Gaussian noise term is replaced with noise drawn from a logistic distribution and coefficients are re-scaled to be one fifth of the size of those used by Friedman, and any $f(\mathbf{X})$ value above 0 is classified as a positive outcome, while $f(\mathbf{X})$ values less than or equal to zero are designated as negative outcomes.  Finally, $f$ is augmented with two hypothetical protected class-control features with known dependencies on the binary outcome to allow for discrimination testing. The simulated data is generated to have eight input features, twelve after numeric encoding of categorical features, and a binary outcome, two class-control features, and 100,000 rows. The simulated data is then split into a training and test set, with 80,000 and 20,000 observations, respectively.  Within the training set, a 5 fold cross validation indicator is used for training all models. For an exact specification of the simulated data, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e., the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contains ten input features and the binary outcome, \textit{high-priced}. The data is split into a training set with 160,338 loans and a marker for 5 fold cross validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (LTV ratio, DTI ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider. See Appendix \ref{a_sec:mort} for general information regarding HMDA data.

@@

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints, with respect to each $X_j$ and $\mathbf{y}$ independently. The MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, and each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$.

\begin{equation}
\begin{aligned}\label{eq:gbm}
g^{\text{MGBM}}(\mathbf{x}) &= \sum_{b=1}^B T_b\left(\mathbf{x}; \Theta^\text{mono}_b\right)
\end{aligned}
\end{equation}

\noindent As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, $\mathbf{y}$, the predictions of all subsequently trained trees in the in $g^{\text{MGBM}}$, $g^{\text{MGBM}}_{b-1}(\mathbf{X})$, and the $b$-th tree splits, $T_b(\mathbf{x}^{(i)};\Theta^\text{mono}_b)$, in a numeric error function (e.g., squared error, Huber error), $l$, and a regularization term that penalizes complexity in the current tree, $\Omega(T_b)$.

Herein, two $g^\text{MGBM}$ models are trained.  One on the simulated data and one on the mortgage data. In both cases, positive and negative monotonic constraints for each $X_j$ are selected using domain knowledge, random grid search is used to determine other hyperparameters, and five-fold cross validation and test partitions are used for model assessment.  See Appendices \ref{a_ssec:not} and \ref{a_ssec:mgbm} for details pertaining to MGBM training. For exact parameterization of the two $g^\text{MGBM}$ models, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e., GAMs which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. XNNs are composed of a global bias term, $\mu_0$, $K$ individually specified neural networks, $n_k$ with scale parameters $\gamma_k$, and the inputs to each $n_k$ are themselves a linear combination of modeling inputs, $\sum_j\beta_{k,j}x_j$.

\begin{equation}
\begin{aligned}
\label{eq:xnn}
g^{\text{XNN}}(\mathbf{x}) = \mu_0 + \sum_{k=0}^{K-1}\gamma_k n_k(\sum^{J=\mathcal{P}-1}_{j=0}\beta_{k, j}x_j)
\end{aligned}
\end{equation}

Here, each $g^\text{XNN}$ is trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in five training folds and in a test data partition. $L_1$ regularization is applied to both the projection and combination layers to induce a sparse and interpretable model, where each $n_k$ subnetwork and corresponding combination layer $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

@@

PD plots are a widely-used method for describing and plotting the average predictions of a complex model $g$ across some partition of data $\mathbf{X}$ for some interesting input feature $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ for a single instance $\mathbf{x} \in \mathcal{X}$ \cite{ice_plots}. PD and ICE can be overlaid in the same plot to compensate for known weaknesses of PD (e.g., inaccuracy in the presence of strong interactions and correlations \cite{ice_plots}, \cite{ale_plot}), to identify interactions modeled by $g$, and to create a holistic global and local portrait of the predictions for some $g$ and $X_j$ \cite{ice_plots}. For details regarding the calculation of one-dimensional PD and ICE, see the software resources in Subsection \ref{ssec:soft} and Appendices \ref{a_ssec:not} and \ref{a_ssec:pd_ice}.

@@

\subsection{Intersectional and Non-static Problems in Machine Learning}

@@

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints, with respect to each $X_j$ and $\mathbf{y}$ independently. The MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, and each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$.

\begin{equation}
\begin{aligned}\label{eq:gbm}
g^{\text{MGBM}}(\mathbf{x}) &= \sum_{b=1}^B T_b\left(\mathbf{x}; \Theta^\text{mono}_b\right)
\end{aligned}
\end{equation}

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e., GAMs which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. XNNs are composed of a global bias term, $\mu_0$, $K$ individually specified neural networks, $n_k$ with scale parameters $\gamma_k$, and the inputs to each $n_k$ are themselves a linear combination of modeling inputs, $\sum_j\beta_{k,j}x_j$.

\begin{equation}
\begin{aligned}
\label{eq:xnn}
g^{\text{XNN}}(\mathbf{x}) = \mu_0 + \sum_{k=0}^{K-1}\gamma_k n_k(\sum^{J=\mathcal{P}-1}_{j=0}\beta_{k, j}x_j)
\end{aligned}
\end{equation}

>>>>

The simulated data (see Subsection \ref{ssec:sim_data}) is based on the well-known Friedman datasets. Its known feature importance and discrimination characteristics are used to gauge the validity of interpretable modeling, post-hoc explanation, and discrimination testing techniques \cite{friedman1979tree}, \cite{friedman1991multivariate}. The mortgage data (see Subsection \ref{ssec:mort_data}) is sourced from the Home Mortgage Disclosure Act (HMDA) database and is a fairly realistic data source for demonstrating the template workflow.\footnote{See: \href{https://www.consumerfinance.gov/data-research/hmda/}{Mortgage data (HMDA)}.} To provide a sense of accuracy differences, performance of more interpretable constrained ML models and less interpretable unconstrained ML models is compared on simulated data and collected mortgage data. Because the unconstrained ML models, gradient boosting machines (GBMs) (e.g. \cite{friedman2001greedy}, \cite{esl}) and artificial neural networks (ANNs) (e.g. \cite{recht2011hogwild}, \cite{hinton2012improving}, \cite{sutskever2013importance}, \cite{zeiler2012adadelta}), do not exhibit convincing accuracy benefits on the simulated or mortgage data and can also present previously delineated risks, further explanation and discrimination analyses are applied only to the constrained, interpretable ML models \cite{please_stop}, \cite{fair_washing}, \cite{scaffolding}. Here, MGBMs\footnote{As implemented in \href{https://xgboost.readthedocs.io/en/latest/tutorials/monotonic.html}{\texttt{XGBoost}} or \href{https://github.com/h2oai/h2o-3/blob/master/h2o-py/demos/H2O_tutorial_gbm_monotonicity.ipynb}{\texttt{h2o}}.} (See subsection \ref{ssec:mgbm}) and XNNs (see Subsection \ref{ssec:xnn}, \cite{wf_xnn} \cite{yang2019enhancing}) will serve as those more interpretable models for subsequent explanatory and discrimination analyses. MGBM and XNN interpretable model architectures are selected for the example workflow because they are straightforward variants of popular unconstrained ML models. If practitioners are working with GBM and ANN models, it should be relatively uncomplicated to also evaluate the constrained versions of these models.

The same can be said of the selected explanation methods and discrimination tests. Due to their post-hoc nature, they can often be shoe-horned into existing ML workflows and pipelines. Presented explanation techniques include PD and ICE (see Subsection \ref{ssec:pd_ice}) and Shapley values (see Subsection \ref{ssec:shap}) \cite{esl}, \cite{ice_plots}, \cite{shapley}, \cite{tree_shap}. PD, ICE, and Shapley values provide direct, global, and local summaries and descriptions of constrained models without resorting to the use of intermediary and approximate surrogate models. Discussed discrimination testing methods (see Subsection \ref{ssec:di}) include AIR, ME, SMD, and confusion matrix metrics \cite{feldman2015certifying}, \cite{cohen1988statistical}, \cite{cohen1992power}.\footnote{Part 1607 - Uniform Guidelines on Employee Selection Procedures (1978) \textsection 1607.4.} Accuracy and other confusion matrix metrics are also reported by demographic segment \cite{zafar2017fairness}. All outlined materials and methods are implemented in open source Python code, and are made available on GitHub (see Subsection \ref{ssec:soft}).

@@

Simulated data is created based on a signal-generating function, $f$, first proposed in \citet{friedman1979tree} and extended in \citet{friedman1991multivariate}:

@@

\noindent where $\mathbf{X}_{\text{Friedman},j}$ are random uniform features in $[0,1]$. In Friedman's texts, a Gaussian noise term was added to create a continuous output feature for testing spline regression methodologies. In this manuscript, the signal generating function and input features are modified in several ways.  Two binary features, a categorical feature with five discrete levels, and a bias term are introduced into $f$ to add a degree of complexity that may more closely mimic real-world settings. 
%When generating the data, random draws for each observation are constructed such that the features can be correlated with each other, some highly so. For example, X1 and X3 have a Pearson correlation coefficient equal to 0.58.
For binary classification analysis, the Gaussian noise term is replaced with noise drawn from a logistic distribution and coefficients are re-scaled to be one fifth of the size of those used by Friedman, and any $f(\mathbf{X})$ value above 0 is classified as a positive outcome, while $f(\mathbf{X})$ values less than or equal to zero are designated as negative outcomes.  Finally, $f$ is augmented with two hypothetical protected class-control features with known dependencies on the binary outcome to allow for discrimination testing. The simulated data is generated to have eight input features, twelve after numeric encoding of categorical features, and a binary outcome, two class-control features, and 100,000 rows. The simulated data is then split into a training and test set, with 80,000 and 20,000 observations, respectively.  Within the training set, a 5-fold cross-validation indicator is used for training all models. For an exact specification of the simulated data, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

The mortgage dataset analyzed here is a random sample of consumer-anonymized loans from the HDMA database. These loans are a subset of all originated mortgage loans in the 2018 HMDA data that were chosen to represent a relatively comparable group of consumer mortgages.  A selection of features is used to predict whether a loan is \textit{high-priced}, i.e., the annual percentage rate (APR) charged was 150 basis points (1.5\%) or more above a survey-based estimate of other similar loans offered around the time of the given loan.  After data cleaning and preprocessing to encode categorical features and create missing markers, the mortgage data contains ten input features and the binary outcome, \textit{high-priced}. The data is split into a training set with 160,338 loans and a marker for 5-fold cross-validation and a test set containing 39,662 loans. While lenders would almost certainly use more information than the selected features to determine whether to offer and originate a high-priced loan, the selected input features (LTV ratio, DTI ratio, property value, loan amount, introductory interest rate, customer income, etc.) are likely to be some of the most influential factors that a lender would consider. See Appendix \ref{a_sec:mort} for general information regarding HMDA data.

@@

MGBMs constrain typical GBM training to consider only tree splits that obey user-defined positive and negative monotonicity constraints, with respect to each input feature, $X_j$, and a target feature, $\mathbf{y}$, independently. An MGBM remains an additive combination of $B$ trees trained by gradient boosting, $T_b$, and each tree learns a set of splitting rules that respect monotonicity constraints,  $\Theta^\text{mono}_b$. For some row of data, $\mathbf{x}$, a trained MGBM model, $g^{\text{MGBM}}$, takes the form:

\begin{equation}
\begin{aligned}\label{eq:gbm}
g^{\text{MGBM}}(\mathbf{x}) &= \sum_{b=0}^{B-1} T_b\left(\mathbf{x}; \Theta^\text{mono}_b\right)
\end{aligned}
\end{equation}

\noindent As in unconstrained GBM, $\Theta^{\text{mono}}_b$ is selected in a greedy, additive fashion by minimizing a regularized loss function that considers known target labels, the predictions of all subsequently trained trees in $g^{\text{MGBM}}$, and the $b$-th tree splits applied to $\mathbf{x}$, $T_b(\mathbf{x}^{(i)};\Theta^\text{mono}_b)$, in a numeric error function (e.g., squared error, Huber error), and a regularization term that penalizes complexity in the current tree. See Appendices \ref{a_ssec:not} and \ref{a_ssec:mgbm} for details pertaining to MGBM training.

Herein, two $g^\text{MGBM}$ models are trained. One on the simulated data and one on the mortgage data. In both cases, positive and negative monotonic constraints for each $X_j$ are selected using domain knowledge, random grid search is used to determine other hyperparameters, and 5-fold cross-validation and test partitions are used for model assessment. For exact parameterization of the two $g^\text{MGBM}$ models, see the software resources referenced in Subsection \ref{ssec:soft}.

@@

XNNs are an alternative formulation of additive index models in which the ridge functions are neural networks \cite{wf_xnn}. XNNs also bear a strong resemblance to generalized additive models (GAMs) and so-called explainable boosting machines (EBMs or GA\textsuperscript{2}M), i.e., GAMs which consider main effects and a small number of 2-way interactions and may also incorporate boosting into their training \cite{esl}, \cite{ga2m}.  Hence, XNNs enable users to tailor interpretable neural network architectures to a given prediction problem and to visualize model behavior by plotting ridge functions. A trained XNN function, $g^{\text{XNN}}$, applied to some row of data, $\mathbf{x}$, is defined as:

\begin{equation}
\begin{aligned}
\label{eq:xnn}
g^{\text{XNN}}(\mathbf{x}) = \mu_0 + \sum_{k=0}^{K-1}\gamma_k n_k(\sum^{J-1}_{j=0}\beta_{k, j}x_j)
\end{aligned}
\end{equation}

\noindent where $\mu_0$ is a global bias for $K$ individually specified ANN subnetworks, $n_k$, with scale parameters $\gamma_k$. The inputs to each $n_k$ are themselves a linear combination of the $J$ modeling inputs and their associated $\beta_{k,j}$ coefficients in the deepest projection layer.

Here, the two $g^\text{XNN}$ models are trained by mini-batch stochastic gradient descent (SGD) on the simulated data and mortgage data. Each $g^\text{XNN}$ is assessed in 5 training folds and in a test data partition. $L_1$ regularization is applied to both the lowest projection layer and to the $n_k$ subnetworks to induce a sparse and interpretable model, where each $n_k$ and corresponding $\gamma_k$ are ideally associated with an important $X_j$ or combination thereof. The $g^\text{XNN}$ models appear highly sensitive to weight initialization and batch size. Be aware that $g^{XNN}$ model architectures may require manual and judicious feature selection due to long training times. For more details regarding $g^\text{XNN}$ training, see the software resources in Subsection \ref{ssec:soft} and Appendices  \ref{a_ssec:not} and \ref{a_ssec:xnn}.

@@

PD plots are a widely-used method for describing and plotting the estimated average predictions of a complex model $g$ across some partition of data $\mathbf{X}$ for some interesting input feature $X_j$ \cite{esl}. ICE plots are a newer method that describes the local behavior of $g$ for a single instance $\mathbf{x} \in \mathcal{X}$ \cite{ice_plots}. PD and ICE can be overlaid in the same plot to compensate for known weaknesses of PD (e.g., inaccuracy in the presence of strong interactions and correlations \cite{ice_plots}, \cite{ale_plot}), to identify interactions modeled by $g$, and to create a holistic global and local portrait of the predictions for some $g$ and $X_j$ \cite{ice_plots}. For details regarding the calculation of one-dimensional PD and ICE, see the software resources in Subsection \ref{ssec:soft} and Appendices \ref{a_ssec:not} and \ref{a_ssec:pd_ice}.

@@

\subsection{Intersectional and Non-static Risks in Machine Learning}

@@

@@

% XNN image
